R---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

`r options(scipen=6)`

## Loading and preprocessing the data

A straightforward command to read in the activity Data Frame from the provided zip file
```{r}
activity_df <- read.csv(unz("activity.zip", "activity.csv"))
```


Let's look at a histogram of the total number of steps taken each day:

```{r}
hist(activity_df$steps,xlab="Steps",main="Steps Taken Each Day by Frequency")

steps_by_day <- tapply( activity_df$steps ,activity_df$date, sum,na.rm=TRUE)

mean_steps_per_day   <- round(mean(steps_by_day))
median_steps_per_day <- median(steps_by_day)
```

Mean steps per day : `r mean_steps_per_day`

Median steps per day: `r median_steps_per_day`






### What is the average daily activity pattern?

Let's make a time series plot of the 5-minute intervals by the average number of steps taken, averaged across all the days.

```{r}



steps_by_interval <- tapply( activity_df$steps ,activity_df$interval, sum,na.rm=TRUE)
intervals         <- rownames(steps_by_interval)

plot(
    intervals,
    steps_by_interval,
    main =  "Average Steps Taken Per 5-Minute Interval" ,
    xlab = "In`terval",
    ylab = "Steps",
    type = "l"
)


max_steps       <- max(steps_by_interval)
which_max_steps <- intervals[ which.max(steps_by_interval) ]

```
The five-minute interval `r which_max_steps` is the interval with the maximums number of steps averaged across all days with `r max_steps`.


### Imputing missing values

How much missing data values (NA) are there?

```{r}
num_of_nas <- sum(is.na(activity_df$steps))
```
The number of NA values in the dataset is `r num_of_nas` out of `r nrow(activity_df)`.


Let's estimate the missing values replacing each NA value with the average of the value for that particular interval.

```{r}
##Find the mean for every interval
mean_steps_by_interval <- tapply(activity_df$steps,activity_df$interval, function(x) { round(mean(x,na.rm=TRUE)) } )

##loop through the DF , if the value is NA then replace it with the mean value for that interval, else just copy
estimated_steps <- vector()
for(i in seq(along=activity_df$steps)) {
    if( is.na(activity_df$steps[i] )) {
        estimated_steps <- c(estimated_steps,mean_steps_by_interval[ as.character(activity_df[i,"interval" ]) ] )
    } else {
        estimated_steps <- c(estimated_steps,activity_df$steps[i])
    }
}

#make a copy of the DF and replace the steps vector with the new estimated steps
activity_df_estimated <- activity_df
activity_df_estimated$steps <- estimated_steps


```

```{r}

hist(activity_df_estimated$steps,xlab="Steps",main="Steps Taken Each Day by Frequency")
steps_by_day <- tapply( activity_df_estimated$steps ,activity_df_estimated$date, sum,na.rm=TRUE)

estimated_mean_steps_per_day       <- round(mean(steps_by_day))
estimated_median_steps_per_day <- median(steps_by_day)

```

What does the new histogram look like now that the NA values have been replaced?


Mean steps per day : `r estimated_mean_steps_per_day`

Median steps per day: `r estimated_median_steps_per_day`


Note that the new median and mean values are larger than the values before we were estimating values for na.RM.  This makes sense given that we are essentially adding new data in to the frame.


### Are there differences in activity patterns between weekdays and weekends?

Let's use dplyr to add a new factor for weekday/weekend and lattice to plot the weekend/weekday data.


```{r}
library(dplyr)

activity_df_estimated <-
    mutate(activity_df_estimated ,
           weekdays =
               ifelse(
                   weekdays(as.Date(activity_df_estimated$date)) == "Saturday" |
                   weekdays(as.Date(activity_df_estimated$date)) == "Sunday" 
                  ,"Weekend","Weekday")
           )

activity_df_estimated$weekdays <- as.factor(activity_df_estimated$weekdays)

weekdays_average_steps <- aggregate(steps ~ interval+weekdays , activity_df_estimated , mean)

library(lattice)
xyplot( steps ~ interval  | factor(weekdays)
      ,data=weekdays_average_steps,type="l",
       xlab = "Interval",
       ylab = "Number of Steps",
       layout=c(1,2)
       )

```

We can see that there are differences between the weekdays and weekends.
Steps start earlier on the weekdays and spike faster (presumably to get to work/school/etc).

